\documentclass[10pt]{article}
\usepackage{geometry}

% For typesetting urls, like my personal and google code pages.
\usepackage{url}

% Changes the margin sizes to create more room for the content itself:
\pagestyle{empty}
\geometry{letterpaper,tmargin=1in,bmargin=0.6in,lmargin=1.1in,rmargin=1.1in,headheight=0in,headsep=0in,footskip=.3in}

% Using the URW Garamond typeface instead of the default:
\usepackage[T1]{fontenc}
% The line spacing needs to be modified a little for the font:
\linespread{1.05}

% Adds mycrotypography support with pdfLaTeX, resulting in better justification.
\usepackage{microtype}

% Removes some unneeded space:
\setlength{\parindent}{0in}
\setlength{\parskip}{0in}
\setlength{\itemsep}{0in}
\setlength{\topsep}{0in}
\setlength{\tabcolsep}{0in}

% Name and contact information:
\newcommand{\name}{Tikhon Jelvis}
\newcommand{\addr}{1100 Pine St., Menlo Park, CA 94025}
\newcommand{\phone}{(650) 391-3532}
\newcommand{\email}{tikhon@berkeley.edu}
\newcommand{\major}{Electrical Engineering and Computer Sciences}
\newcommand{\university}{University of California, Berkeley}
\newcommand{\degree}{Bachelor of Science}
\newcommand{\website}{\url{inst.eecs.berkeley.edu/~tikhon}}

%--------------------------------------------------%
% New Commands:
% Creates the header of the resume:
\newcommand{\rsheader}[1]{
  \begin{center}
    \fontfamily{phv}
    \selectfont\Huge\scshape#1
  \end{center}
}

% Resume section commands:
\newenvironment{ressection}[1]{
  {\fontfamily{phv}\selectfont\LARGE#1}\vspace{2pt}\\
}{  
\vspace{4pt}
}

\newenvironment{ressubsection}[1]{
  {\fontfamily{phv}\selectfont\Large#1}
  \fontfamily{ppl}\selectfont\\
}
{
}

\newenvironment{reslist}{
  \vspace{-22pt}
  \begin{itemize}
    \setlength{\itemsep}{1pt}
    \setlength{\parskip}{0pt}
    \setlength{\parsep}{0pt}
}{  
  \end{itemize}
  \vspace{-4pt}
}

\begin{document}

% Name followed by a horizontal line:
\rsheader{\name} \vspace{-10pt} \rule{\textwidth}{1pt}

% Contact information:
\vspace{-1pt} {\small\itshape \addr \hfill \phone; \email}

\begin{ressection}{Objective}
  A summer programming internship, preferably with a focus on web
  programming, JavaScript and Ajax technologies.
\end{ressection}

\begin{ressection}{Education}
  \university \\ 
  \degree, \major,
  expected May 2014\\
  G.P.A. 3.891\\
  SAT: 2370, Verbal: 800, Math: 770, Writing: 800
\end{ressection}

\begin{ressection}{Awards}
  \begin{reslist}
  \item National Merit Finalist, 2010
  \item AMC 12 school winner in 2010 with 100.5 points (97.5th
    percentile in the country).
  \item Overall math contest school winner, 2010
  \item Bank of America Certificate for Computer Studies, 2010
  \item Menlo Atherton Rotary Club Subject Award for Chemistry, 2008
    and Computer Science, 2009
  \end{reslist}
\end{ressection}

\begin{ressection}{Skills}
  \begin{ressubsection}{Computer Software}
    \begin{reslist}
    \item Linux (OpenSUSE, Ubuntu)
    \item Windows (XP, Vista, 7)
    \item Emacs and Eclipse
    \item Command-line Bash
    \item SVN
    \end{reslist}
  \end{ressubsection}

  \begin{ressubsection}{Programming}
    \begin{reslist}
    \item Proficient in Java, JavaScript, (x)HTML, CSS and \LaTeX
    \item Familiar with Scheme, elisp, Perl, regular expressions,
      Haskell and C++
    \item Experienced in prototype.js, script.aculo.us, jQuery and
      Java Swing
    \end{reslist}
  \end{ressubsection}
  \begin{ressubsection}{Other}
    \begin{reslist}
    \item Fluent in Russian
    \item Competitive saber fencer
    \end{reslist}
  \end{ressubsection}
\end{ressection}

\begin{ressection}{Experience}
  \begin{ressubsection}{Robotics}
    Worked on programming for FIRST Robotics team 766 (Menlo Atherton
    High School). Using OOP design and Java, implemented a program to
    control the robot. Led a small team including several novice
    programmers. Leveraged a vision-recognition library to find a
    target with the robot's camera.
  \end{ressubsection}

  \begin{ressubsection}{Java Physics Simulation}
    Worked with a group of three other people, including two
    non-programmers, on a Java Swing application for modeling 2D
    collisions between polygons. Resulted in a program with about 8500
    lines of code and over 40 files.
  \end{ressubsection}

  \begin{ressubsection}{Maptac Chess}    
    Developed a JavaScript chess game using prototype.js and
    script.aculo.us. Integrated this with a Chess AI a friend wrote in
    JavaScript. Released the project under the MIT license on Google
    Code. It can be found at
    \url{code.google.com/p/maptac-chess}.
  \end{ressubsection}
\end{ressection}

\rule{\textwidth}{1pt}

\begin{center}
  More details online at: \url{inst.eecs.berkeley.edu/~tikhon}
\end{center}

\end{document}